Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-27 article.tex
Edits and comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\begin{introduction}
Parsing is "the process of analysing a string of symbols according to the rules
of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building 
a tree representing the structure of the parsed input, upon which further 
operations such as interpretation or code generation are then performed.

In addition to recognizing a stream of tokens and building a parse tree, a
real-world parser must recognize faulty input, and accurately report the
location and cause of the problem.  But how -- what generates a parse error,
what information is included in the error, how do error-reporting parsers
compose?

Monad transformers provide a clean and simple answer to these questions.
We'll use them to build a parser combinator library, then we'll use the library
to build a simple parser, and finally we'll add in error reporting.
\end{introduction}

\section{Parser Combinators}
Parser combinators are an excellent approach for building expressive, 
declarative, and easy-to-maintain parsers.  They also benefit from host-language
integration, allowing them to snarf features such as type systems and test 
frameworks.

There are many excellent papers on parser combinators, such
as Wadler's classic paper \cite{wadler} on non-deterministic parser combinators, 
Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
and Leijen's Parsec paper \cite{leijen}.  
The inspiration and and much of the names used in this article are drawn 
from those references.

To get a feel for what parser combinators are and how they're used to build
parsers, let's make a quick pass through a minimal parser combinator library 
so that we can see them in action before we move on to the hard stuff.  We'll 
divide the library into three main pieces of the code:
\begin{itemize}
 \item a Parser datatype
 \item parser-specific combinators
 \item type class instances
\end{itemize}

\subsection{Parser datatype}
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}
This type definition says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.

\subsection{Type class instances}
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

@@

\begin{verbatim}
Woof         :=   Form(+)

Form         :=   Symbol  |  Special  |  Application

Symbol       :=   [a-zA-Z](+)

Special      :=   '{'  ( Define  |  Lambda )  '}'

Define       :=   'define'  Symbol  Form

Lambda       :=   'lambda'  '{'  Symbol(*)  '}'  Form(+)

Application  :=   '('  Form(+)  ')'

Whitespace   :=   \s+

Comment      :=   ';'  (not '\n')(*)
\end{verbatim}

>>>>

\begin{introduction}

\EZY{I rewrote the abstract, let me know if you like it.}

The goal of parsing, in abstract, is to take a string and build a tree
representing the structure of this input according to a formal grammar.
However, real-world parsers must also recognize faulty input and
accurately report the location and cause of the problem.  What generates
a parse error, how is that information included in the error, and how do
error-reporting parsers compose?

Monad transformers provide a clean and simple answer to these questions.
We'll show how monad transformers can be used to build a parser combinator
and then add error reporting in later in a modular way.

\end{introduction}

\section{Parser Combinators}

\EZY{It's probably worth rephrasing what is stated in the abstract here,
for those of us who are in the habit of skipping abstracts; it also helps
build continuity.}

Parser combinators are an excellent approach for building expressive, 
declarative, and easy-to-maintain parsers.  They also benefit from host-language
integration, allowing them to snarf features such as type systems and test 
frameworks.

There are many excellent papers on parser combinators, such
as Wadler's classic paper \cite{wadler} on non-deterministic parser combinators, 
Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
and Leijen's Parsec paper \cite{leijen}.
The inspiration and many of the names used in this article are drawn 
from those references.

To get a feel for what parser combinators are and how they're used to
build parsers, we'll quickly create a minimal parser combinator library to see
the basic principles of their design before adding monad transformers to
the mix.  We'll divide the library into three main pieces of the code:

\begin{itemize}
 \item a Parser datatype
 \item parser-specific combinators
 \item type class instances
\end{itemize}

\subsection{Parser datatype}
\EZY{These sections are very terse.  If the reader is expected to already be familiar with
most of the moving bits of a parsing library, perhaps this section can be boiled down some more.
It is a very curious audience the current presentation targets: a reader who knows how to
implement a parser library ``from scratch'', but doesn't know how to use
monad transformers to build them.  Perhaps the components of this
presentation which are key to understanding how the monad transformer works should be
fleshed out more.}

\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}
This type definition defines Parser to be a function
on a list of tokens, producing a new list of tokens and a result
value, possibly failing.

\subsection{Type class instances}
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
\EZY{Fun fact: it is possible to create parsers which only have an applicative instance,
and are more efficient but less expressive than the style you have here!}
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

@@

\begin{verbatim}
Woof         :=   Form(+)
Form         :=   Symbol  |  Special  |  Application
Symbol       :=   [a-zA-Z](+)
Special      :=   '{'  ( Define  |  Lambda )  '}'
Define       :=   'define'  Symbol  Form
Lambda       :=   'lambda'  '{'  Symbol(*)  '}'  Form(+)
Application  :=   '('  Form(+)  ')'
Whitespace   :=   \s+
Comment      :=   ';'  (not '\n')(*)
\end{verbatim}
